---
output: github_document
bibliography: references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stats19

The goal of stats19 is to ease access, processing and analysis of the UK's official road traffic casualty database.

It builds on previous work including:

- code in the [bikeR](https://github.com/Robinlovelace/bikeR) repo underlying an academic paper on collisions involving cyclists [@lovelace_who_2016]
- functions in [**stplanr**](https://github.com/ropensci/stplanr/blob/master/R/load-stats19.R) for downloading Stats19 data
- updated functions related to the [CyIPT](https://github.com/cyipt/stats19) project.

## Installation

Install the latest version with:

```{r install, results='hide', message=FALSE}
devtools::install_github("ITSLeeds/stats19")
```


<!-- You can install the released version of stats19 from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("stats19") -->
<!-- ``` -->

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
## basic example code
```

## References
