---
output: github_document
bibliography: vignettes/references.bib
---

[![Travis build status](https://travis-ci.org/ITSLeeds/stats19.svg?branch=master)](https://travis-ci.org/ITSLeeds/stats19)
[![codecov](https://codecov.io/gh/ITSLeeds/stats19/branch/master/graph/badge.svg)](https://codecov.io/gh/ITSLeeds/stats19)

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stats19

The goal of **stats19** is to make it easy to work with road crash data.
Specifically it enables access to and processing of the UK's official road traffic casualty database, which is called [STATS19](https://data.gov.uk/dataset/cb7ae6f0-4be6-4935-9277-47e5ce24a11f/road-safety-data).
The name comes from the form used by the police to record car crashes and other incidents resulting in casualties on the roads.

A description of the stats19 data and variables they contain can be found in a [document](http://data.dft.gov.uk/road-accidents-safety-data/Brief-guide-to%20road-accidents-and-safety-data.doc) hosted by the UK's Department for Transport (DfT).

The package builds on previous work including:

- code in the [bikeR](https://github.com/Robinlovelace/bikeR) repo underlying an academic paper on collisions involving cyclists
- functions in [**stplanr**](https://github.com/ropensci/stplanr/blob/master/R/load-stats19.R) for downloading Stats19 data
- updated functions related to the [CyIPT](https://github.com/cyipt/stats19) project

## Installation

Install and attach the latest version with:

```{r install, eval=FALSE}
devtools::install_github("ITSLeeds/stats19")
```

```{r attach}
library(stats19)
```

<!-- You can install the released version of stats19 from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("stats19") -->
<!-- ``` -->

## Data download

**stats19** enables download of raw stats19 data with `dl_*` functions.
The following code chunk, for example, downloads and unzips a .zip file containing Stats19 data from 2017:

```{r example}
dl_stats19(years = 2017, type = "Accidents")
```

Currently, these files are downloaded to a default location of "tempdir" which is a platform independent "safe" location to download the data in. Once downloaded, they are unzipped under original DfT file names. The function prints out the location and final file name(s) of unzipped files(s) as shown above.

Data files from other years can be downloaded in an interactive manner, providing just the year for example, would result in options presented to you:

```{r, eval=TRUE}
dl_stats19(years = 2017)
```

## Reading-in data

Downloaded data can then be read-in as follows (assuming the data download went OK):

```{r}
d17 = "dftRoadSafetyData_Accidents_2017"
dl_stats19(file_name = paste0(d17, ".zip"))
crashes_2017_raw = read_accidents(years = 2017, filename = "Acc.csv")
crashes_2017 = format_accidents(crashes_2017_raw)
```

What just happened?
We read-in data on all road crashes recorded by the police in 2017 across Great Britain.
`read_accidents()` imports the 'raw' Stats19 data without cleaning messy column names or re-categorising the outputs.
`format_accidents()` does this work, automating the process of matching column names with variable names and labels in a [`.xls` file](http://data.dft.gov.uk/road-accidents-safety-data/Road-Accident-Safety-Data-Guide.xls) provided by the DfT.
This means `crashes_2017` is much more usable than `crashes_2017_raw`, as shown below, which shows three records and some key variables in the messy and clean datasets:

```{r}
key_patt = "severity|speed|light|human"
key_vars = grep(key_patt, x = names(crashes_2017_raw), ignore.case = TRUE)
random_n = sample(x = nrow(crashes_2017_raw), size = 3)
crashes_2017_raw[random_n, key_vars]
crashes_2017[random_n, key_vars]
```

<!-- More data can be read-in as follows: -->

```{r, eval=FALSE, echo=FALSE}
# old code to be up-dated
d14 = "Stats19_Data_2005-2014"
crashes_2005_2014 = read_accidents(data_dir = d14)
crashes_2005_2014_f = format_stats19_2005_2014_ac(crashes_2005_2014)
d15 = "RoadSafetyData_2015"
crashes_2015 = read_accidents(data_dir = d15, filename = "Accidents_2015.csv")
crashes_2015_f = format_stats19_2015_ac(crashes_2015)
d16 = "dftRoadSafety_Accidents_2016"
crashes_2016 = read_accidents(data_dir = d16, filename = "dftRoadSafety_Accidents_2016.csv")
crashes_2016_f = format_stats19_2016_ac(crashes_2016)
all_crashes = rbind(crashes_2015_f, crashes_2016_f, crashes_2017_f)
table(ac$Accident_Severity)
```
