
WARNING: this does not work on Windows.

This function generates the data object `stats19_schema` in a reproducible way
using DfT's schema definition (see function [dl_schema()]).

The function also generates `stats19_variables` (see the function's source code
for details).

# Load stats19 schema

```{r}
read_schema = function(
  data_dir = tempdir(),
  filename = "Road-Accident-Safety-Data-Guide.xls",
  sheet = NULL
) {

  file_path = file.path(data_dir, filename)
  if(!file.exists(file_path)) {
    dl_schema()
  }
  if(is.null(sheet)) {
    export_variables = readxl::read_xls(path = file_path,
                                        sheet = 2,
                                        skip = 2)
    export_variables_collisions = tibble::tibble(
      table = collision,
      variable = export_variables$`Accident Circumstances`
    )
    export_variables_vehicles =  tibble::tibble(
      table = "vehicles",
      variable = export_variables$Vehicle
    )
    export_variables_casualties =  tibble::tibble(
      table = "casualties",
      variable = export_variables$Casualty
    )
    export_variables_long = rbind(
      export_variables_collisions,
      export_variables_casualties,
      export_variables_vehicles
    )
    stats19_variables = stats::na.omit(export_variables_long)
    stats19_variables$type = stats19_vtype(stats19_variables$variable)

    # add variable linking to names in formatted data
    names_acc = names(accidents_sample)
    names_veh = names(vehicles_sample)
    names_cas = names(casualties_sample)
    names_all = c(names_acc, names_veh, names_cas)

    variables_lower = schema_to_variable(stats19_variables$variable)
    # test result
    # variables_lower[!variables_lower %in% names_all]
    # names_all[!names_all %in% variables_lower]
    stats19_variables$column_name = variables_lower
    # head(stats19_variables)

    #' export result: usethis::use_data(stats19_variables, overwrite = TRUE)

    sel_character = stats19_variables$type == "character"
    character_vars = stats19_variables$variable[sel_character]
    character_vars = stats19_vname_switch(character_vars)

    schema_list = lapply(
      X = seq_along(character_vars),
      FUN = function(i) {
        x = readxl::read_xls(path = file_path, sheet = character_vars[i])
        names(x) = c("code", "label")
        x
      }
    )

    stats19_schema = do.call(what = rbind, args = schema_list)
    n_categories = vapply(schema_list, nrow, FUN.VALUE = integer(1))
    stats19_schema$variable = rep(character_vars, n_categories)

    character_cols = stats19_variables$column_name[sel_character]
    stats19_schema$variable_formatted = rep(character_cols, n_categories)

  } else {
    stats19_schema = readxl::read_xls(path = file_path, sheet = sheet)
  }
  stats19_schema
}
```

# Download schema

```{r}
dl_schema = function(data_dir = tempdir()) {
u = paste0(
  "https://data.dft.gov.uk/road-accidents-safety-data/",
  "Road-Accident-Safety-Data-Guide.xls"
)
destfile = file.path(data_dir, "Road-Accident-Safety-Data-Guide.xls")
utils::download.file(u, destfile = destfile)
# download and unzip the data if it's not present
}
```

# Function to covert schema names

```{r}
schema_to_variable = function(x) {
  x = format_column_names(x)
  x = gsub(pattern = " ", replacement = "_", x = x)
  x = gsub(pattern = "_null_if_not_known", replacement = "", x)
  x = gsub(pattern = "_dd/mm/yyyy|_hh:mm", replacement = "", x)
  x = gsub(pattern = "highway_authority___ons_code", replacement = "highway", x)
  x = gsub(pattern = "lower_super_ouput_area", replacement = "lsoa", x)
  x = gsub(pattern = "_england_&_wales_only", replacement = "", x)
  x = gsub(pattern = "_cc", replacement = "", x)
  x = gsub(pattern = "vehicle_propulsion_code", replacement = "propulsion_code", x)
  x = gsub(pattern = "pedestrian_road_maintenance_worker_from_2011",
           replacement = "pedestrian_road_maintenance_worker", x)
  x = gsub(pattern = "engine_capacity", replacement = "engine_capacity_cc", x)
  x = gsub(pattern = "age_of_vehicle_manufacture", replacement = "age_of_vehicle", x)
  x
}
```

# Return type of variable of stats19 data

```{r}
stats19_vtype = function(x) {
  variable_types = rep("character", length(x))
  sel_numeric = grepl(pattern = "Number|Speed|Age*.of|Capacity", x = x)
  variable_types[sel_numeric] = "numeric"
  sel_date = grepl(pattern = "^Date", x = x)
  variable_types[sel_date] = "date"
  sel_time = grepl(pattern = "^Time", x = x)
  variable_types[sel_time] = "time"
  sel_location = grepl(pattern = "^Location|Longi|Lati", x = x)
  variable_types[sel_location] = "location"
  #' remove other variables with no lookup: no weather ?!
  sel_other = grepl(
    pattern = paste0(
      "Did|Lower|Accident*.Ind|Reference|Restricted|",
      "Leaving|Hit|Age*.Band*.of*.D|Driver*.H"
    ),
    x = x
  )
  variable_types[sel_other] = "other"
  variable_types
}

stats19_vname_switch = function(x) {
  x = gsub(pattern = " Authority - ONS code", "", x = x)
  x = gsub(
    pattern = "Pedestrian Crossing-Human Control",
    "Ped Cross - Human",
    x = x
    )
  x = gsub(
    pattern = "Pedestrian Crossing-Physical Facilities",
    "Ped Cross - Physical",
    x = x
    )
  x = gsub(pattern = "r Conditions", "r", x = x)
  x = gsub(pattern = "e Conditions", "e", x = x)
  x = gsub(pattern = " Area|or ", "", x = x)
  x = gsub(pattern = "Age Band of Casualty", "Age Band", x = x)
  x = gsub(pattern = "Pedestrian", "Ped", x = x)
  x = gsub(pattern = "Bus Coach Passenger", "Bus Passenger", x = x)
  x = gsub(pattern = " \\(From 2011\\)", "", x = x)
  x = gsub(pattern = "Casualty Home Type", "Home Area Type", x = x)
  x = gsub(pattern = "Casualty IMD Decile", "IMD Decile", x = x)
  x = gsub(pattern = "Driver IMD Decile", "IMD Decile", x = x)
  x = gsub(pattern = "Journey Purpose of Driver", "Journey Purpose", x = x)
  x
}
```

Informal test:

```{r}
variable_types = stats19_vtype(stats19_variables$variable)
names(variable_types) = stats19_variables$variable
variable_types
x = names(get_stats19(year = 2022, type = collision, ask = FALSE))
n = stats19_vtype(x)
names(n) = x
n
```

The data bundled with the package can then be (re-)generated with

```{r}
stats19_schema = read_schema()
View(stats19_schema)
```
