---
title: "Introducing stats19"
author: 
  - "Layik Hama and Robin Lovelace"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{stats19-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

**stats19** is an R package that enables access to and processing of Great Britain's official road traffic casualty database, [STATS19](https://data.gov.uk/dataset/cb7ae6f0-4be6-4935-9277-47e5ce24a11f/road-safety-data). 

The name comes from a form by the police in the UK called [STATS19](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/775149/Operational_Metrics_Manual.pdf) (capital letters). A short googling for this blogpost did not result in finding the origin of the name but there is at least STATS20 which is guidelines to officers to fill in STATS19 form. That is why th Department for Transport (DfT) also names the dataset [STATS19](https://data.gov.uk/dataset/cb7ae6f0-4be6-4935-9277-47e5ce24a11f/road-safety-data).

The importance of road safety and informed decision making based on crash data, in our view, cannot be overrated. Two numbers from a strategy document by the UK government (2011) is worth mentioning here: 

> The economic welfare costs [of road collisions] are estimated at around £16 billion a year while insurance payouts for motoring claims alone are now over £12 billion a year.

## The dataset

**stats19** package is the result of earlier work by Dr Robin Lovelace (maintainer) and Dr Malcom Morgan of Institute of Transport Studies at the University of Leeds. They had used the dataset before work started on the package. 

There are three different CSV files released by the DfT: `accidents`, `vehicles` and `casualties`. There is a schema covering these tables but a good amount of work is needed to understand it and be able to use it.

The annual statistics are not released in a systematic way either. For example, there are separate files for each of the above tables for certain years but not for all of 1979 - 2017 or 2018 now. The largest chunk is the 1979 - 2004 file ([link](http://data.dft.gov.uk/road-accidents-safety-data/Stats19-Data1979-2004.zip)), unzipped they are:   
```sh
721M Apr  3  2013 Accidents7904.csv
344M Apr  3  2013 Casualty7904.csv
688M Apr  3  2013 Vehicles7904.csv
# total 1.753 GB data
```

## **stats19**

With those introductions out of the way, lets see what the package can do and how it empower academics, policy makers, road safety campaigners and other R programmers to have access to a large dataset from 1979 - now. 

```{r, eval=FALSE, message=FALSE}
install.packages("stats19")
# the dev version is used for this blog post
remotes::install_github("ropensci/stats19")
```

```{r}
library(stats19)
library(kableExtra)
```

The easiest way to get STATS19 data is with `get_stats19()`. This function takes 2 main arguments, `year` and `type`. The year can be any year between 1979 and 2017.

```{r dl2017-accidents, message=FALSE}
crashes_2017 = get_stats19(year = 2017, type = "Accidents", ask = FALSE)
nrow(crashes_2017)
```

What just happened?
We just downloaded, cleaned and read-in data on all road crashes recorded by the police in 2017 across Great Britain. We can explore the `crashes_2017` object a little more:

```{r crashes_2017-explore}
column_names = names(crashes_2017)
length(column_names)
head(column_names)
class(crashes_2017)
kable(head(crashes_2017[, column_names[1:5]]))
```

We can then convert the object ino

The package contains the names of all "zip" files released by the DfT and hosted on Amazon servers to download. These file names have been included in the package and can be found under `file_names` variable name. for example:

```{r file_names}
stats19::file_names$dftRoadSafetyData_Vehicles_2017.zip
```

## Creating geographic crash data

An important feature of STATS19 data is that the "accidents" table contains geographic coordinates.
These are provided at ~10m resolution in the UK's official coordinate reference system (the Ordnance Survey National Grid, EPSG code 27700).
**stats19** converts the non-geographic tables created by `format_accidents()` into the geographic data form of the [`sf` package](https://cran.r-project.org/package=sf) with the function `format_sf()` as follows:

```{r format-crashes-sf}
crashes_sf = format_sf(crashes_2017)
```

An example of an administrative zone dataset of relevance to STATS19 data is the boundaries of police forces in England, which is provided in the packaged dataset `police_boundaries`.
The following code chunk demonstrates the kind of spatial operations that can be performed on geographic STATS19 data, by counting and plotting the number of fatalities per police force:

```{r nfatalities, message=FALSE}
library(sf)
library(dplyr)
crashes_sf %>% 
  filter(accident_severity == "Fatal") %>% 
  select(n_fatalities = accident_index) %>% 
  aggregate(by = police_boundaries, FUN = length) %>% 
  plot()
```

```{r ukboundaries}
west_yorkshire =
  police_boundaries[police_boundaries$pfa16nm == "West Yorkshire", ]
```


```{r crashes-west_yorkshire}
crashes_wy = crashes_sf[west_yorkshire, ]
nrow(crashes_wy) # which is 3.36%
```

## Combining the tables
We can combine the three sets of tables to analyse the data further. Lets read the datasets first:

```{r dl2017-vehcas, message=FALSE}
#crashes_2017 = get_stats19(year = 2017, type = "Accidents", ask = FALSE)
casualties_2017 = get_stats19(year = 2017, type = "Casualties", ask = FALSE)
nrow(casualties_2017)
vehicles_2017 = get_stats19(year = 2017, type = "Vehicles", ask = FALSE)
nrow(vehicles_2017)
```

Lets now read in casualties that took place in West Yorkshire (using `crashes_wy` object above), and count the number of casualties by severity for each crash:

```{r table-join, message = FALSE}
library(tidyr)
library(dplyr)
sel = casualties_2017$accident_index %in% crashes_wy$accident_index
casualties_wy = casualties_2017[sel, ]
cas_types = casualties_wy %>% 
  select(accident_index, casualty_type) %>% 
  group_by(accident_index) %>% 
  summarise(
    Total = n(),
    walking = sum(casualty_type == "Pedestrian"),
    cycling = sum(casualty_type == "Cyclist"),
    passenger = sum(casualty_type == "Car occupant")
    ) 
cj = left_join(crashes_wy, cas_types)
```

What just happened? 

We found the subset of casualties that took place in West Yorkshire with reference to the `accident_index` variable in the `accidents` table.
Then we used the **dplyr** function `summarise()`, to find the number of people who were in a car, cycling, and walking when they were injured.
This new casualty dataset is joined onto the `crashes_wy` dataset.
The result is a spatial (`sf`) data frame of crashes in West Yorkshire, with columns counting how many road users of different types were hurt.
The joined data has additional variables:

```{r table-join-examples}
base::setdiff(names(cj), names(crashes_wy))
```

As a simple spatial plot, we can map all the crashes that have happened in West Yorkshire in 2017, with the colour related to the total number of people hurt in each crash.
Placing this plot next to a map of West Yorkshire provides context:

In addition to the `Total` number of people hurt/killed, `cj` contains a column for each type of casualty (cyclist, car occupant, etc.), and a number corresponding to the number of each type hurt in each crash.
It also contains the `geometry` column from `crashes_sf`.
In other words, joins allow the casualties and vehicles tables to be geo-referenced.
We can then explore the spatial distribution of different casualty types.
The following figure, for example, shows the spatial distribution of pedestrians and car passengers hurt in car crashes across West Yorkshire in 2017:

```{r sfplot, fig.show='hold', out.width="100%", fig.cap="Spatial distribution of serious and fatal crashes in West Yorkshire in 2017, for cycling, walking, being a car passenger and other modes of travel. Colour is related to the speed limit where the crash happened (red is faster) and size is proportional to the total number of people hurt in each crash.", fig.width=9, fig.height=7}
library(ggplot2)
crashes_types = cj %>% 
  filter(accident_severity != "Slight") %>% 
  mutate(type = case_when(
    walking > 0 ~ "Walking",
    cycling > 0 ~ "Cycling",
    passenger > 0 ~ "Passenger",
    TRUE ~ "Other"
  ))
ggplot(crashes_types, aes(size = Total, colour = speed_limit)) +
  geom_sf(show.legend = "point", alpha = 0.3) +
  facet_grid(vars(type), vars(accident_severity)) +
  scale_size(
    breaks = c(1:3, 12),
    labels = c(1:2, "3+", 12)
    ) +
  scale_color_gradientn(colours = c("blue", "yellow", "red")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

```{r crashes-map, fig.show='hold', out.width="100%", fig.cap="Spatial distribution of all crashes in 2017 in West Yorkshire.", fig.width=9, fig.height=7}
library(leaflet)
# convert to lon lat CRS
wy_lonlat = st_transform(crashes_wy, crs = 4326)
c = st_bbox(wy_lonlat)
leaflet(data = wy_lonlat$geometry, height = "280px") %>%
  addProviderTiles(provider = providers$OpenStreetMap) %>%
  fitBounds(lat1 = c$ymin[[1]], lat2 = c$ymax[[1]],
            lng1 = c$xmin[[1]], lng2 = c$xmax[[1]]) %>% addCircleMarkers(radius = 0.5, color = "blue", weight = 1) 
```

## Conclusion

You have access to a reliable and official road safety dataset which is cleaned up and ready to be consumed. We hope this package can be used by campaigners, academics and other interestd parties to make the roads of the UK and the rest of the world a safe place for all of us.
