---
title: "DuckDB and Parquet engines"
author: "R Lovelace"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DuckDB and Parquet engines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

## Introduction

The **stats19** package supports two engines for reading and filtering data: `readr` (the default) and `duckdb`. 
The `duckdb` engine is particularly useful for working with large datasets, as it can filter data at the database level before loading it into R, significantly reducing memory usage and processing time.

This vignette demonstrates that the two engines produce equivalent results, then
shows how to cache the data as Parquet for faster repeated reads.

## Load the package

```{r}
library(stats19)
```

## Equivalence test with 2024 data

We will use the 2024 collision data to compare the two engines.

```{r, message=FALSE, warning=FALSE}
# Read with readr (default)
col_readr = get_stats19(year = 2024, type = "collision", engine = "readr", silent = TRUE)

# Read with duckdb
col_duckdb = get_stats19(year = 2024, type = "collision", engine = "duckdb", silent = TRUE)
```

### Comparing record counts and fatal counts

Both engines should return the same number of records and the same number of fatal collisions.

```{r}
nrow(col_readr)
nrow(col_duckdb)

# Check fatal counts
sum(col_readr$collision_severity == "Fatal", na.rm = TRUE)
sum(col_duckdb$collision_severity == "Fatal", na.rm = TRUE)
```

### Using waldo for deep comparison

We can use the `waldo` package to check for any differences between the first 10 records.

```{r}
if (requireNamespace("waldo", quietly = TRUE)) {
  waldo::compare(head(col_readr, 10), head(col_duckdb, 10))
}
```

## Working with large datasets (1979-latest)

When working with the full historical dataset (from 1979 onwards), the `duckdb` engine is highly recommended. The following code demonstrates how to use the `duckdb` engine with a `where` clause to filter the data efficiently.

```{r, eval=FALSE}
# This chunk is not evaluated because it requires downloading ~1.5GB of data
# and can take several minutes to run with the readr engine.

# Download and read all collisions since 1979, but only keep those with speed_limit = 30
crashes_30mph = get_stats19(year = 1979, type = "collision", 
                           engine = "duckdb", 
                           where = "speed_limit = 30")
```

The `duckdb` engine can be more than 50 times faster than the `readr` engine when performing such filtered reads on large files.

## Caching the data as Parquet

The `duckdb` engine reads the `.csv` files directly. That is fast, but it still
parses text every time you call `get_stats19()`. Converting the files once to
Parquet makes repeated reads faster again: Parquet files are compressed, typed
and columnar, so a query only touches the columns it needs.

### Where Parquet files live

Parquet files are written to the directory returned by `get_parquet_directory()`,
which is inside your data directory by default. Set the
`STATS19_PARQUET_DIRECTORY` environment variable, or call
`set_parquet_directory()`, to keep them somewhere else:

```{r, eval=FALSE}
# Persist the setting by adding STATS19_PARQUET_DIRECTORY=/path/to/parquet to
# your .Renviron file, which you can open with usethis::edit_r_environ()
set_parquet_directory("/data/silver/dft/stats19")
get_parquet_directory()
```

### Converting CSVs to Parquet

`stats19_to_parquet()` converts the raw files already in your data directory.
There is no need to download them first: `dl_stats19()` does that.

```{r, eval=FALSE}
library(stats19)

# Convert every collisions CSV in the data directory into one Parquet file
stats19_to_parquet(type = "collision")

# Casualties, vehicles and collisions in one call
stats19_to_parquet(type = "all")
```

Useful arguments:

* `years` converts only the years you ask for, for example `years = 2021:2025`.
* `partition_by` writes a Hive-partitioned directory instead of a single file,
  for example `partition_by = "collision_year"`. Partitioning is worth it when
  you usually query a single year.
* `compression` is `"zstd"` by default. `"snappy"` is slightly faster to write
  and larger on disk.
* `max_mem_gb` caps DuckDB's memory use (4 GB by default) and `temp_dir` sets
  where it spills to disk, which matters when converting the 1979-latest files.
* `overwrite = FALSE` leaves an existing Parquet file alone.

The function returns the path it wrote, invisibly.

### Reading from the cache

Set `engine = "parquet"` to read from the cache. If a Parquet file is missing,
or does not cover the years you asked for, stats19 warns and falls back to the
CSV files it has on disk.

```{r, eval=FALSE}
col = get_stats19(year = 2024, type = "collision", engine = "parquet", silent = TRUE)
```

To keep the data in DuckDB and only pull the results into R, use
`output_format = "duckdb"`. This returns a lazy `dbplyr` table, so `dplyr` verbs
are translated into SQL and no rows are transferred until you call `collect()`:

```{r, eval=FALSE}
library(dplyr)

tbl = get_stats19(year = 2021:2025, type = "collision", output_format = "duckdb")

tbl |>
  filter(collision_severity == "Fatal") |>
  count(collision_year) |>
  collect()
```

Two things to know about `output_format = "duckdb"`:

* It needs the `dbplyr` package, and it is only available for data that lives in
  DuckDB, so a year covered by no CSV and no Parquet file will come back empty.
* The DuckDB connection stays open so that the lazy table keeps working, and
  stats19 will not close it for you. Close it when you are done, or you will
  accumulate open connections:

```{r, eval=FALSE}
DBI::dbDisconnect(dbplyr::remote_con(tbl), shutdown = TRUE)
```

### Keep the cache up to date

Parquet files are not rebuilt when new CSVs arrive. After downloading a new
year, convert again, otherwise queries for that year find a cache that stops
short of it and fall back to reading CSV:

```{r, eval=FALSE}
dl_stats19(year = 2026, type = "collision")
# rebuild so the cache includes 2026
stats19_to_parquet(type = "collision")
```

The same applies when DfT reissues an existing year: the cached copy keeps the
old values until you rebuild it.
