---
title: "Introducing stats19"
author: 
  - "R Lovelace, M Morgan, L Hama and M Padgham"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{stats19-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**stats19** enables access to and processing of the UK's official road traffic casualty database, [STATS19](https://data.gov.uk/dataset/cb7ae6f0-4be6-4935-9277-47e5ce24a11f/road-safety-data). 
A description of variables in the database can be found in a [document](http://data.dft.gov.uk/road-accidents-safety-data/Brief-guide-to%20road-accidents-and-safety-data.doc) provided by the UK's Department for Transport (DfT).
The datasets are collectively called STATS19 after the form used to report them, which can be found [here](http://docs.adrn.ac.uk/888043/mrdoc/pdf/888043_stats19-road-accident-injury-statistics-report-form.pdf). 
This vignette focusses on how to use the **stats19** package to work with STATS19 data.

**Note**: The Department for Transport refers to "accidents", but "crashes" is a more appropriate term, as emphasised in the "crash not accident" arguments of road safety advocacy groups such as [RoadPeace](http://www.roadpeace.org/take-action/crash-not-accident/).
We use the term "accidents" only in reference to nomenclature within the data as provided.

The development version is hosted on [GitHub](https://github.com/ITSLeeds/stats19) and can be installed and attached as follows:

```{r, eval=FALSE}
devtools::install_github("ITSLeeds/stats19")
```

```{r}
library(stats19)
```

**stats19** has three main types of functions, corresponding with the three stages of data processing:

- **Download**: This single `dl_stats19()` function accepts `year`, `type` and `filename` arguments to make it easy to find the right file for your research question.

- **Read**: STATS19 data is provided in a particular format that benefits from being read-in with pre-specified column types. This is taken care of with `read_*()` functions providing access to the 3 main tables in STATS19 data:

    - `read_accidents()` reads-in the crash data (which has one row per incident)
    - `read_casualties()` reads-in the casualty data (which has one row per person injured or killed)
    - `read_vehicles()` reads-in the vehicles table, which contains information on the vehicles involved in the crashes (and has one row per vehicle)

- **Format**: Each of the `read_*()` functions has a `format` parameter which, when `TRUE`, adds labels to the tables. The raw data provided by the DfT contains only integers. 
Running `read_*(..., format = TRUE)` converts these integer values to the corresponding character variables for each of the three tables.
For example, `read_accidents(format = TRUE)` converts values in the `accident_severity` column from `1`, `2` and `3` to `Slight`, `Serious` and `Fatal`.
To read-in raw data without formatting, set `format = FALSE`.

Multiple functions are needed for each step because of the structure of STATS19 data, which are divided into 3 tables, as described in an article that analyses 2005-2015 STATS19 data with reproducible R code by Christoph Freier ([2018](https://kawameicha.gitlab.io/inCodeWeTrust/2018/01/10-years-and-3.3-million-accidented-vehicles-in-the-uk/)):

1. "accident circumstances, with details about location, severity, weather, etc; 
2. casualties, referencing knowledge about the victims; and
3. vehicles, which contains more information about the vehicle type and maneuvers, as well the some information about the driver."

The data can be downloaded for many years.
Datasets since 1979 are broadly consistent, meaning that STATS19 data represents a very rich historic geographic, as stated in the DfT's road casualties report in [2017](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/744077/reported-road-casualties-annual-report-2017.pdf):

> The current set of definitions and detail of information 
goes back to 1979, providing a long period for comparison.

## Download STATS19 data

The function `dl_stats19` downloads individual files, and takes three parameters.
Supplying a `file_name` is interpreted to mean that the user is aware of what to download and the other two parameters will be ignored.
You can also use `years` and `type` to "search" through the file names, which are stored in a lazy-loaded datset called `stats19::file_names`.

You can find out the names of files that can be downloaded with `names(stats19::file_names)`, an example of which is shown below:

```{r}
stats19::file_names$dftRoadSafetyData_Vehicles_2017.zip
```

To see how `file_names` was created, see `?file_names`.
The following code chunk downloads and unzips a .zip file containing STATS19 data from 2017:

```{r dl2017-accidents}
dl_stats19(year = 2017, type = "Accidents")
```

Data files from other years can be selected interactively.
Just providing a year, for example, presents the user with multiple options (from `file_names`), illustrated below:

```{r dl2017-all, eval=FALSE}
dl_stats19(year = 2017)
```
```
Multiple matches. Which do you want to download?

1: dftRoadSafetyData_Vehicles_2017.zip
2: dftRoadSafetyData_Casualties_2017.zip
3: dftRoadSafetyData_Accidents_2017.zip

Selection: 
Enter an item from the menu, or 0 to exit
```

When R is running interactively, you can select which of the 3 matching files to download:
those relating to vehicles, casualties or accidents in 2017.
You can also search for file names, as illustrated below:

```{r dl2017-search}
d17 = "dftRoadSafetyData_Accidents_2017"
dl_stats19(file_name = paste0(d17, ".zip"))
```

Note that this is the same file (containing the 'accidents' table from 2017) downloaded previously with the command `dl_stats19(year = 2017, type = "Accidents")`.
If the file is already present, `dl_stats19()` will skip the download process.

## Read STATS19 data

In a similar approach to the download section before, we can read files downloaded using a `data_dir` location of the file and the `filename` to read.
The code below will download the `dftRoadSafetyData_Accidents_2017.zip` file from the DfT servers and read its content.
Files are saved by default in `tempdir()`, but this can be overridden to ensure permanent storage in a user-defined location.

```{r dl2017-read}
d17 = "dftRoadSafetyData_Accidents_2017"
dl_stats19(file_name = paste0(d17, ".zip"))
crashes_2017_raw = read_accidents(data_dir = tempdir(),
                                  year = 2017,
                                  filename = "Acc.csv")
```

**stats19** imports data with `readr::read_csv()` which results in a 'tibble' object: a data frame with more user-friendly printing and a few other features.

```{r crashes2017-class}
class(crashes_2017_raw)
dim(crashes_2017_raw)
```

## Format STATS19 data

A [`.xls` file](http://data.dft.gov.uk/road-accidents-safety-data/Road-Accident-Safety-Data-Guide.xls) provided by the DfT defines the column names for the datasets provided.
The packaged datasets `stats19_variables` and `stats19_schema` provide summary information about the contents of this data guide.
These contain the full variable names in the guide (`stats19_variables`) and a complete lookup table relating integer values to the `.csv` files provided by the DfT and their labels (`stats19_schema`).
The first rows of each dataset are shown below:

```{r variables-and-schema}
stats19_variables
stats19_schema
```

The code that generated these small datasets can be found in their help pages (accessed with `?stats19_variables` and `?stats19_schema` respectively).
`stats19_schema` is used internally to automate the process of formatting the downloaded `.csv` files.
Column names are formatted by the function `format_column_names()`, as illustrated below:

```{r format-col-names}
format_column_names(stats19_variables$variable[1:3])
```

Previous approaches to data formatting `STATS19` data involved hard-coding results.
This more automated approach to data cleaning is more consistent and fail-safe.
Three functions, `format_accidents()`, `format_vehicles()` and
`format_casualties()` do the data formatting on the respective data frames, as
illustrated below:

```{r format-main}
crashes_2017 = format_accidents(crashes_2017_raw)

# vehicle data for 2017
dl_stats19(year = 2017, type = "vehicles")
vehicles_2017_raw = read_vehicles(year = 2017)
vehicles_2017 = format_vehicles(vehicles_2017_raw)

# casualties data for 2017
dl_stats19(year = 2017, type = "casualties")
casualties_2017_raw = read_casualties(year = 2017)
casualties_2017 = format_casualties(casualties_2017_raw)
```
The package is designed to automatic this two-step `read_*` and `format_*`
process by defaulting in all cases to `data_year = read_*(year, format = TRUE)`,
so that the single `read_*` functions return by default the formatted data. The
two-step process may nevertheless be important for reference to the official
nomenclature and values as provided by the DfT.

A summary of the outputs for each of the three tables is shown below.

```{r summarise-stats19}
summarise_stats19 = function(x) {
  data.frame(row.names = 1:length(x),
    name = substr(names(x), 1, 19),
    class = sapply(x, class),
    n_unique = sapply(x, function(v) length(unique(v))),
    first_label = sapply(x, function(v) substr(unique(v)[1], 1, 9)),
    second_label = sapply(x, function(v) substr(unique(v)[2], 1, 9))
  )
}
```

```{r summarise-crashes}
knitr::kable(summarise_stats19(crashes_2017), 
             caption = "Summary of formatted crash data.")
```

```{r summarise-vehicles}
knitr::kable(summarise_stats19(vehicles_2017), 
             caption = "Summary of formatted vehicles data.")
```

```{r summarise-casualties}
knitr::kable(summarise_stats19(casualties_2017), 
             caption = "Summary of formatted casualty data.")
```

For testing and other purposes, a sample from the accidents table is provided in the package.
A few columns from the two-row sample is shown below:

```{r, echo=FALSE, results='asis'}
key_patt = "severity|speed|light|human"
key_vars = grep(key_patt, x = names(stats19::accidents_sample_raw), ignore.case = TRUE)
knitr::kable(stats19::accidents_sample_raw[, key_vars])
```

<!-- ## Analysis: an example of pedestrians hurt in Leeds -->


## Further work

There is much potential to extend the package, beyond downloading, reading and formatting STATS19 data.
The greatest potential is to provide functions that will help with analysis of STATS19 data, to help with road safety research.
Much academic research has been done using the data, a few examples of which are highlighted below to demonstrate the wide potential for further work.

- Research exploring the effectiveness of road safety policies such as speed limits. An example in this area is @grundy_effect_2009, who found that areas with 20mph speed limits were safer. This raises the question: can the same result be repeated using reproducible methods? Does the finding hold for more recent 20 mph zones? Is the recent finding of DfT's ([2018](https://www.gov.uk/government/publications/20-mph-speed-limits-on-roads)) research, that 20 mph zones alone do not reduce crash rates, supported by reproducible analysis? What are the factors that make speed limits more or less effective [see @sarkar_street_2018 for example]?
- Research into weather as a contributing factor to road traffic casualties [e.g. @edwards_relationship_1998]. This raises the question: could matching crash data from the **stats19** data with historic weather data from other R packages help advance knowledge in this area?
- Assesment of crash rates normalised by estimated exposure rates (risk). An example of this type of research by an author of the package found substantial spatial variation in the number of cyclist casualties across West Yorkshire [@lovelace_who_2016]. This raises the questions: are similar spatial differences found in other regions? What are the factors leading to relatively high and low rates of different types of crash? 

The broader point is that the **stats19** could help road safety research, by making open access data on road crashes more accessible to researchers worldwide.
By easing the data download and cleaning stages of research, it could also encourage reproducible analysis in the field.

There is great potential to add value to and gain insight from the data by joining the datasets with open data, for example from the Consumer Data Research Centre ([CDRC](www.cdrc.ac.uk), which funded this research), OpenStreetMap and the UK's Ordnance Survey.
If you have any suggestions on priorities for these future directions of (hopefully safe) travel, please get in touch on at [github.com/ITSLeeds/stats19/issues](https://github.com/ITSLeeds/stats19/issues).

## References
