---
title: "Introduction to R for road safety: an introduction to R and practical exercises"
subtitle: "Practical exercises based on UK data from the stats19 package, developed by the Institute for Transport Studies, University of Leeds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{stats19-training}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
author: Robin Lovelace, Malcolm Morgan & Andrea Gilardi
bibliography:
  - references.bib
  - packages.bib
---


# Introduction

This document provides information, code and, vitally, exercises to test and improve your R skills.
It starts with introductory R skills that will be of use in any domain but the focus is on R for Road Safety, in support of a [2 day course](https://www.racfoundation.org/introduction-to-r-for-road-safety).
The course is based on open road crash records from the **stats19** package [@lovelace_stats19_2019].
Code and data supporting the content can be found in the package's GitHub repo at [github.com/ropensci/stats19](https://github.com/ropensci/stats19/).
The '[issue tracker](https://github.com/ropensci/stats19/issues)' associated with that repo is a good place to ask questions about the course. 

## Prerequisites

If you are not experienced with R, it is strongly advised that you read-up on and more importantly **test out** R and RStudio before attempting analyse road crash data with R.
See the `stats19-training-setup` vignette at https://docs.ropensci.org/stats19/articles/stats19-training-setup.html for guidance on getting started with R, RStudio and installing R packages.

Additional course pre-requisites are outlined in the [stats19-training-setup](https://docs.ropensci.org/stats19/articles/stats19-training-setup.html) hosted at [docs.ropensci.org/stats19](https://docs.ropensci.org/stats19).
It makes use of a number of packages which can be installed with `install.packages()` and loaded as follows:

```{r, message=FALSE, warning=FALSE, eval=FALSE}
library(pct)      # access travel data from DfT-funded PCT project 
library(sf)       # spatial vector data classes
library(stats19)  # get stats19 data
library(stplanr)  # transport planning tools
library(tidyverse)# packages for 'data science'
library(tmap)     # interactive maps
```

You should type, run and ensure you understand each line of code in this document.


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "50%"
)
```

```{r pkgs, warning=FALSE, echo=FALSE}
pkgs = c(
  "sf",          # spatial data package
  "stats19",     # downloads and formats open stats19 crash data
  "dplyr",       # a package data manipulation, part of the tidyverse
  "tmap"         # for making maps
)
```

```{r cite, echo=FALSE}
knitr::write_bib(x = pkgs, "packages.bib")
```

```{r, eval=FALSE}
remotes::install_cran(pkgs)
# remotes::install_github("ITSLeeds/pct")
```

The workshop is based on a practical handout that can be found [here](https://github.com/ropensci/stats19/releases/download/1.1.0/stats-19-exercises.pdf).

# Introduction

This vignette supports workshops on road safety analysis with R.
It was initially developed for a 2 day [Introduction to R for Road Safety](https://www.racfoundation.org/introduction-to-r-for-road-safety) course funded by the [RAC Foundation](https://www.racfoundation.org/), a charity that supports road safety research.

The course is based on open access road crash data from the UK, which is provided in an accessible form to R users through the R package **stats19** [@lovelace_stats19_2019].
However, the content should be of use for anyone working with road crash data that has (at a minimum):

- A timestamp
- A location (or address that can be geocoded)
- Attribute data, such as severity of crash, mode of vehicles involved etc.


# Agenda

You can work through this tutorial in your own time.
If done as part of a course, the agenda at [`stats19-training-setup`](https://docs.ropensci.org/stats19/articles/stats19-training-setup.html)

The preliminary agenda is as follows:

## Day 1: An introduction to R and RStudio for spatial and temporal data

09:00-09:30 Arrival and set-up

09:30-11:00 Introduction to the course and software

- Introduction to R + coding [video](https://youtu.be/7oyiPBjLAWY?t=357)/[links](https://github.com/jennybc/code-smells-and-feels)
- R installation questions/debugging
- How to use RStudio (practical in groups of 2)
- R classes and working with data frames (CC)

**Break**

11:15-12:30 Working with temporal data

- Time classes
- Filtering by time of crash
- Aggregating over time
- Forecasting crashes over time

**Lunch**

13:30-15:00 Working with spatial data

- Spatial data in R
- Context: spatial ecosystem (see section [1.4 of Geocomputation with R - package ecosystem](https://geocompr.robinlovelace.net/intro.html#rs-spatial-ecosystem))
- [Exercises](https://geocompr.robinlovelace.net/attr.html#exercises-1): Section 6 of the handout

- Further reading: [Section 3.2 to 3.2.2](https://geocompr.robinlovelace.net/attr.html#vector-attribute-manipulation) of handouts

**Break**
  
15:15-15:30 Talk on Road Safety 1

15:30-16:15 Practical - Applying the methods to stats19 data - a taster

- How to access data with **stats19**
- Key **stats19** functions
- Excercises: analysing road crash data on the Isle of Wight

16:15-16:30 Talk on Road Safety 2

## Day 2 road safety analysis with R

09:30-11:00 Point pattern analysis

- Visualising data with tmap
- Spatial and temporal subsetting
- Aggregation

11:15-12:30 Road network data

- Desire lines: using origin-destination data
- Downloading road network data from OSM
- Buffers on road networks

**Lunch**

13:30-15:00 Analysing crash data on road network

**Break**
  
15:15-15:30: Talk on Road Safety 3

15:30-16:30 Applying the methods to your own data

# References
