---
title: "An introduction to road safety analysis with R: setup notes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{stats19-training-setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "50%"
)
```

If you are not experienced with R, it is strongly advised that you read-up on and more importantly **test out** R and RStudio before attempting analyse road crash data with R.

To read up on R, we recommend reading Chapter 1 Getting Started with Data in R of the online book Statistical Inference via Data Science, which can be found here: https://moderndive.netlify.com/1-getting-started.html

Reading sections 1.1 to 1.3 of that book and trying a few of the examples are considered **essential prerequisites**, unless you are already experienced with R.

Optionally, if you want a more interactive learning environment, you can try getting started with the free [DataCamp](https://www.datacamp.com/courses/free-introduction-to-r) course.
Other good resources can be found at [education.rstudio.com/learn](https://education.rstudio.com/learn/beginner/).

And for more information on how R can be used for transport research, the Transportation chapter of Geocomputation with R is a good place to start: https://geocompr.robinlovelace.net/transport.html

**Your computer should also have the necessary software installed.**

To ensure your computer is ready for the course, you should have a recent (3.6.0 or later) version of R or RStudio installed. You should have installed packages stats19, tidyverse and a few others shown below.
To check you have the necessary packages installed, try running the following lines of R code in the RStudio console (this will install the packages you need if they are not already installed):

```{r, eval=FALSE}
install.packages("remotes")
pkgs = c(
  "pct",         # package for getting travel data in the UK
  "sf",          # spatial data package
  "stats19",     # downloads and formats open stats19 crash data
  "stplanr",     # for working with origin-destination and route data
  "tidyverse",   # a package for user friendly data science
  "tmap"         # for making maps
)
remotes::install_cran(pkgs)
# remotes::install_github("ITSLeeds/pct")
```

To test your computer is ready to work with road crash data in R, try running the following commands from RStudio (which should result in the map below):

 <!-- method for helping people set up their computers. Type this single line into the console and follow the instructions.  -->

```{r message=FALSE, eval=FALSE}
library(stats19)
library(tidyverse)
library(tmap) # installed alongside mapview
crashes = get_stats19(year = 2017, type = "ac")
crashes_iow = crashes %>% 
  filter(local_authority_district == "Isle of Wight") %>% 
  format_sf()
  
# basic plot
plot(crashes_iow)
```

You should see results like those shown in the map here: https://github.com/ropensci/stats19/issues/105

If you cannot create that map by running the code above before the course, get in touch with us, e.g. by writing a comment under that github issue page (Note: You will need a github account). 
